---
title: "Final project 2019"
output:
  pdf_document: default
  html_document: null
  df_print: paged
---

The attached \verb,hurrican356.csv, included the track data of 356 hurricanes in  the North Atlantic area since 1989. For all the storms, their location (longitude \& latitude) and maximum wind speed were recorded every 6 hours. The data includes the following variables 

1. **ID**:  ID of the hurricans
2. **Season**: In which \textbf{year} the hurricane occurred 
3. **Month**: In which \textbf{month} the hurricane occurred 
4. **Nature**:  Nature of the hurricane 
  + ET: Extra Tropical
  + DS: Disturbance
  + NR: Not Rated
  + SS: Sub Tropical
  + TS: Tropical Storm
5. **time**: dates and time of the record  
6. **Latitude** and **Longitude**:  The location of  a hurricane check point 
7. **Wind.kt**  Maximum wind speed (in Knot) at each check point 


Researchers want to develop a model to predict the  hurricane trajectory.



##  Load  and plot the hurrican data

```{r eval=FALSE, include=TRUE}
library(ggplot2)
dt= read.csv("/Users/yw2148/Dropbox/Teaching/Teaching-computing/My Teaching Files/big data computing/1_ggplot/hurrican356.csv")
ggplot(data=dt, aes(x = Longitude, y = Latitude)) + 
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE)
library(data.table)
dt <- as.data.table(dt)
summary(dt)
```
Overlay the hurrican data in the world map
```{r eval=FALSE, include=TRUE}
library(maps)
map <- ggplot(data = dt, aes(x = Longitude, y = Latitude)) + 
  geom_polygon(data = map_data(map = 'world'), aes(x = long, y = lat, group = group))
map +
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE, alpha = 0.75) + 
  ggtitle(paste0("Atlantic Windstorm mean knot"))
```
Additional Plots
```{r eval=FALSE, include=TRUE}
map <- ggplot(dt, aes(x = Longitude, y = Latitude, group = ID)) + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               fill = "gray25", colour = "gray10", size = 0.2) + 
  geom_path(data = dt, aes(group = ID, colour = Wind.kt), size = 0.5) + 
  xlim(-138, -20) + ylim(3, 55) + 
  labs(x = "", y = "", colour = "Wind \n(knots)") + 
  theme(panel.background = element_rect(fill = "gray10", colour = "gray30"),
        axis.text.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

seasonrange <- paste(range(dt[, Season]), collapse=" - ")

map + ggtitle(paste("Atlantic named Windstorm Trajectories (", 
                     seasonrange, ")\n")) 
```
Show hurricance tracks by month
```{r eval=FALSE, include=TRUE}
mapMonth <- map + facet_wrap(~ Month) +
  ggtitle(paste("Atlantic named Windstorm Trajectories by Month (", 
                 seasonrange, ")\n")) 
mapMonth
```

# Problem 1  (60 points)


Let $t$ be time (in hours) since a hurricane began, and For each hurrican $i$, we denote 
$\{ Y_{i,1}(t), Y_{i,2}(t), Y_{i,3}(t)\},j=1,2,3$ be the latitude, longitude, and wind speed at time $t$. We consider the following model 

$$Y_{i,j}(t+6) =  \mu_{i,j}(t) +\rho_jY_{i,j}(t) + \epsilon_{i,j}(t)$$
where $\mu_{i,j}(t)$ is the funtional mean, and the errors $(\epsilon_{i,1}(t),\epsilon_{i,2}(t),\epsilon_{i,3}(t))$ follows a multivariate normal distributions with mean zero and covariance matrix $\Sigma$, independent across $t$. We further assume that the mean functions $\mu_{i,j}(t)$ can be written as
$$\mu_{i,j}(t) =  \beta_{0,j} +x_{i,1}(t)\beta_{1,j} +
+x_{i,2} \beta_{2,j} + x_{i,3}\beta_{3,j} +
\sum_{k=1}^3\beta_{3+k,j}\Delta_{i,k}(t-6)
$$ where
$x_{i,1}(t)$, ranging from 0 to 365, is the day of year at time $t$, $x_{i,2}$ is the calenda year of the hurrican, and $x_{i,3}$ is the type of hurrican, 
and
$$\Delta_{i,k}(t-6) = Y_{i,k}(t) -Y_{i,k}(t-6),k=1,2,3$$
are the change of latitude, longitude, and wind speed
between $t-6$ and $t$.

\paragraph{Prior distribution} We assume the prior distribution of 
$\boldsymbol{\beta}=(\beta_{k,j})_{k=0,...,6, j=1,2,3}$, $\pi(\boldsymbol{\beta})$ is jointly normal with mean 0 and variance $diag(1, p)$.

$\pi(\rho_j)$ follows a trucated normal  $N_{[0,1]}(0.5, 1/5)$ 
 
 $\pi(\Sigma^{-1})$ follows a $Wishart (3, diag(0.1,3))$


\paragraph{Your to-do-list:}
\begin{enumerate}
\item Randomly select 80\% hurricanes and develop an MCMC algorithm to estiamte the posterior mean of the model  parameters.

\item Apply your model to track the remaining 20\% hurricans, and evaluate how well your model could predict and track these hurricanes. 

\item Write a summary to report your findings. 
\end{enumerate}


# Problem 2 (40 points)
The attached \verb,hurricanoutcome2.csv,  recorded the damages and death caused by 46 hurricanes in the U.S.  The variables include 


1. **ID**:  ID of the hurricans
2. **Season**: In which \textbf{year} the hurricane occurred 
3. **Month**: In which \textbf{month} the hurricane occurred 
4. **Nature**:  Nature of the hurricane 
  + ET: Extra Tropical
  + DS: Disturbance
  + NR: Not Rated
  + SS: Sub Tropical
  + TS: Tropical Storm
5. **Damage**: Financial loss (in Billion U.S. dollars) caused by hurricanes
6. **Deaths**: Number of death caused by hurricanes
7. **Maxspeed**: Maximum recorded wind speed of the hurricane
8. **Meanspeed**: average  wind speed of the hurricane
9. **Maxpressure**: Maximum recorded central pressure of the hurricane
10. **Meanpressure**: average central pressure of the hurricane
11. **Hours**: Duration of the hurricane in hours 
12. **Total.Pop**:  Total affected population   
13. **Percent.Poor**: \% affected population that reside in poor countres (i.e. GDP per Capita $<=$ 10,000)  
14. **Percent.USA**:  \% affected population that reside in the United States   



Please propose a model to investigate which characteristics of the hurricanes are associated with damage and deaths, and how well they could predict  the hurricane induced damage and deaths.   Propose estimation method/algorithm, interpret your estimated models, and evaluate the prediction power of your estimated models. 








  
